Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-06-18 Manuscript/TEM_draft.tex
Correction methods

%HEADER
\documentclass[12pt,letterpaper]{article}

%packages
% NC: Just to flag here how I'll comment on LaTeX files. I'll use % and then NC. Of course in git diff it'll all show up as an addition.
% NC: You could add all the packages etc to a template .sty file to save writing it all up here. Rich's ms.sty that I provided on DropBox is a good example
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{verbatim}
\usepackage[english]{babel}
\usepackage{pifont}
\usepackage{color}
\usepackage{setspace}
\usepackage{lscape}
\usepackage{indentfirst}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
%\usepackage{nag}
\usepackage{natbib}
%\usepackage{bibtex}
\usepackage{float}
\usepackage{latexsym}
%\usepackage{hyperref} 
\usepackage{url}
%\usepackage{html}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
%\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{amstext} % Add and remove packages as necessary for your manuscript.

@@

\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
% We don't use a special title page; the author information is entered like any other text.
% FOOTNOTES: We don't allow them in the manuscript, except in tables. Don't include any footnotes in the text.
%IMPORTANT:
%Note on display in the text file: sentences or sub-parts of sentences (e.g. enumerations) are delimited by a newline for clarity.
%However, true new lines (i.e. in the pdf) are delimited by one or more empty lines.
%We don't use a heading for the introduction. The nature of this section is implied by its position following the abstract and preceding the first major section heading.

% NC: You should delete this extraneous info from the Syst Biol template to tidy everything up a bit

@@

%For a colour version of the draft to a search>replace '-BW'>'' 
% %\modulolinenumbers[1] 	% Line numbering on every line

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Missing data and topology in total evidence matrices %RH = Running Head

@@

%Title
\noindent{\Large \bf Effect of missing data in matrices containing living and fossil taxa on topological accuracy}
%Alternative punchier title: "Bayesian total evidence methods fail to recover correct topologies from matrices containing both living and fossil taxa"
%Alternative punchier title: "Maximum Likelihood outperforms Bayesian methods for inferring topology from total evidence matrices containing both living and fossil taxa"

% NC: Not sure about any of these. I actually think I prefer your talk title. So something like: The effects of missing data on the ability of total evidence approaches to recover correct tree topologies. No rush to decide. But needs total evidence in there somewhere.
\bigskip

%Authors
\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, other authors $^3$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;\\
$^3$Somewhere else }\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}
% NC: It's not called the Republic of, just Ireland!

\newpage
\begin{abstract}
% NC: I'm not going to give this much work as it'll change before we submit anyway. Note though that and abstract must NOT just repeat stuff in the intro. Needs rephrasing at the very least.
Living species represent less than 1\% of all species that have ever lived. % NC: I was chatting to Andy Purvis about this the other day and he said lots of palaeo folk hate that stat. So let's not start the abstract with it!
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both fossil and living taxa must be included in macroevolutionary studies.
One approach, the total evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both fossil and living taxa at the tips.
Although the total evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from total evidence matrices.
Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix:
(1) the proportion of living taxa with no morphological data,
(2) the amount of missing data in the fossil record, and
(3) the overall number of morphological characters in the matrix.
We find that, in a Bayesian framework, difficulties in recovering a stable topology are mainly driven by the missing data in the molecular part of the matrix (for which fossil taxa have no data).
In a Maximum Likelihood framework, however, topology is not directly affected by missing data \textit{per se}, but by the number of morphological characters shared among the taxa.
Therefore, the two main drivers of incorrect topologies are the overall number of morphological characters and the number of living species with no morphological data.

Our results suggest that, in order to use total evidence approaches, one should reduce the missing data in the morphological part of the matrix for living species and use a Maximum Likelihood framework to fix the topology prior to the overall Bayesian phylogenetic inference process.
%More results!
\end{abstract}

\noindent (Keywords: missing data, total evidence, Bayesian, Maximum Likelihood, topology)\\

@@

Although most species that have ever lived are now extinct % NC: This avoids giving the dodgy stat of 1%
 \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species (e.g., \citealp{meredithimpacts2011}). %or is citing of your own papers in the too "douchebagy" for the second sentence?
% NC: Yeah be careful with self citing. Fine if it's a good fit, but it's totally not here!!! A better citation would be a big biodiversity study, e.g. the bird phylogeny paper.
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}. This has led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013}). %Not a nice sentence

Up to now, three main approaches have been used to place both living and fossil taxa % NC: BE consistent. Either living and fossil, or fossil and living.
into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). % NC: Not sure about bit in brackets
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty (but see \citealp{spencerefficacy2013}). Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence dates % NC: Need to check the exact phrase needed here. Is it range of occurrence? Occurrence dates? etc.
 are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013}.
However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa. %There might be a philosophico-semantic distinction to do here. Is TEM a method or an approach? I'll say that it is an approach because it's more like an idea (put every thing together) than a way to do (how to put everything together). It makes more sense to me compared to a real method e.g. a probabilistic method such as ML where the method states precisely how to do stuff.
% NC: Tricky. Don't think it's too important though. 
% NC: I do think Total Evidence needs capital letters though, like ML and Bayesian
This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{eernissetaxonomic1993}. Total Evidence methods have been successfully applied to empirical data \citep{pyrondivergence2011,ronquista2012,schragocombining2013}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies. However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data. In particular it requires morphological data from both living and fossil taxa, both of which are known to be scarce. % NC: Really? I know fossil data is known to be scarce, but living data? I think people would be surprised. And they are surprised. We were surprised! Needs a citation either way.
Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}. % Trevor: Missing data also influences support values that trend to decrease

% NC: You need to take a bit of time to learn about singular and plural grammar in English. It's something you always do wrong. e.g. The Total Evidence method *uses* this and *its* parameters are as follows. Versus: Total Evidence methods *use* this and *their* parameters are as follows. Sive and I won't be around to correct your English forever.

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013}.
Missing molecular data has been seen by some authors as an issue because it can decrease phylogenetic signal (i.e. % define phylogenetic signal here) 
in some parts of the tree, especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered % what do you mean by highly-covered?
genes (i.e. approximatively half of the genes; \citealt{rouresite-specific2011});
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and
(iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}.
Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013}.
Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies \citep{sansomfossilization2013}.
However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point.

@@

% NC: Somewhere here you need to flag that you're only looking at topologies and briefly explain why.

Here we use simulations to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. The molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}. The effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}, therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix:
\begin{enumerate}
\item the proportion of living taxa with no morphological data;
\item the amount of missing data in the fossil taxa; and
\item the overall number of morphological characters for both living and fossil taxa in the matrix.
\end{enumerate}
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect real-life biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases (citation?), and 
% OK I'm not sure what the "real-life" reason for three is.
%This paragraph is crap
% NC: I think it's fine!

We find that when using a Maximum Likelihood approach, as missing data increases, the likelihood of recovering the correct tree topology decreases. However, even with no missing data, Total Evidence matrices dramatically reduce the performance of Bayesian methods for inferring tree topology. We propose that this drastic difference between Bayesian and Maximum Likelihood methods is due to a flattening of the likelihood landscape caused by the unavoidable amount of missing molecular data for fossil taxa in a Total Evidence matrix. We make suggestions for how best to deal with this issue when inferring phylogenies from Total Evidence matrices.

% NC: You don't want to tell the whole story here. Just a results summary and a quick line of interpretation

%This problem leads to an inability of Bayesian chains to pick up the sub and near optimal topologies leading to high variance in the posterior tree distribution and really low support in the consensus tree.
%These results should encourage authors in not using Bayesian inference to explore the topological landscape of their phylogeny.
%Instead, we suggest that authors should first find the Maximum Likelihood tree and use it's topology as a fix prior in their Bayesian inference.
%We also propose a new method for comparing Bayesian posterior tree distribution's topology available in a wonderful R package. %Or am I just dreaming a bit too much here?

% NC: Dreaming a bit much I think. There's a discussion to be had here about whether it's really a *new* method and whether you really want to make an R package per se. We definitely want to share the code

@@

\newpage
\section{Methods} %no material needed?
% NC: Depends on the journal. See what Syst Biol asks for as section titles.
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}). %NC: does Syst Biol use Fig. or Figure?
\begin{enumerate}
\item{Generating the matrix} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the "true" tree; Table ~\ref{Tab_glossary}) and used it to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix; Table ~\ref{Tab_glossary}).
\item{Removing data} \label{step:remove_data} \\
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of missing living taxa ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$) % NC: make sure to use the same descriptions here and in the introduction 
(the resulting matrices are called hereafter "missing-data" matrices; Table ~\ref{Tab_glossary}).
\item{Building phylogenies} \label{step:build_phylo} \\
We built phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the "best" tree; Table ~\ref{Tab_glossary}) and multiple trees inferred from matrices with missing morphological data (hereafter called the "missing-data" trees; Table ~\ref{Tab_glossary}). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies} \label{step:compare_topo} \\
We compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies.
\end{enumerate}
We repeated steps 1-4 50 times. %that's because I had 3 cluster accounts available... If it's a too odd number, no worries, I'll just remove the last chain.
% NC: Yeah 51 is odd. Let's just use 50.
A list of all the terms used in this paper is available in Table ~\ref{Tab_glossary}.

\begin{figure}
\centering
\includegraphics[keepaspectratio=true]{Figures/TEM_Fig_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 124 "missing-data" matrices.
% NC: I thought we had 124 missing data matrices... I just worked this out - one of the "missing data" matrices is 0%,0%,0% so = complete matrix.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated steps 1-4 50 times.}
\label{Fig_Outline}
\end{figure}

% NC: One tiny edit to the figure. I think the ML then Bayesian story is nice, so maybe change that part of the figure so it's ML then Bayesian. Also let's show this figure to a couple of people to see if they get it.

@@

% NC: I think this section is suffering a bit from "here is everything I know about tree comparison metrics" rather than explaining what is important for your analysis. You may be able to cut some more stuff out.

>>>>

 \documentclass[12pt,letterpaper]{article}

%packages
% NC: Just to flag here how I'll comment on LaTeX files. I'll use % and then NC. Of course in git diff it'll all show up as an addition.
% NC: You could add all the packages etc to a template .sty file to save writing it all up here. Rich's ms.sty that I provided on DropBox is a good example

% TG: Just a general note for the file outline: I write every sentence of a same paragraph on a new line, no real reason for that appart that I can then see when the sentence is way to long. Also, I kept the big hidden headers ("%----") that you don't seem to like, without them it totally wrecks my head to know how much do I still have to scroll...

\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
%\usepackage{nag}
\usepackage{natbib}
%\usepackage{bibtex}
\usepackage{float}
\usepackage{latexsym}
%\usepackage{hyperref} 
\usepackage{url}
%\usepackage{html}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
%\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{amstext} % Add and remove packages as necessary for your manuscript.

@@

\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib

%---------------------------------------------
%
%       START
%
%---------------------------------------------

@@

%For a colour version of the draft to a search>replace '-BW'>'' 
%\modulolinenumbers[1]     % Line numbering on every line

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Missing data and topology in Total Evidence matrices %RH = Running Head

@@

%Title
\noindent{\Large \bf Effect of missing data in matrices containing living and fossil taxa on topological accuracy}
%Alternative punchier title: "Bayesian Total Evidence methods fail to recover correct topologies from matrices containing both living and fossil taxa"
%Alternative punchier title: "Maximum Likelihood outperforms Bayesian methods for inferring topology from Total Evidence matrices containing both living and fossil taxa"
% NC: Not sure about any of these. I actually think I prefer your talk title. So something like: The effects of missing data on the ability of Total Evidence approaches to recover correct tree topologies. No rush to decide. But needs Total Evidence in there somewhere.

\bigskip

%Authors
\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, other authors $^3$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;\\
$^3$Somewhere else }\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}
% NC: I'm not going to give this much work as it'll change before we submit anyway. Note though that and abstract must NOT just repeat stuff in the intro. Needs rephrasing at the very least.
Living species represent less than 1\% of all species that have ever lived. % NC: I was chatting to Andy Purvis about this the other day and he said lots of palaeo folk hate that stat. So let's not start the abstract with it!
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both living and fossil taxa must be included in macroevolutionary studies.
One approach, the Total Evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both living and fossil taxa at the tips.
Although the Total Evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix:
(1) the proportion of living taxa with no morphological data,
(2) the amount of missing data in the fossil record, and
(3) the overall number of morphological characters in the matrix.
We find that, in a Bayesian framework, difficulties in recovering a stable topology are mainly driven by the missing data in the molecular part of the matrix (for which fossil taxa have no data).
In a Maximum Likelihood framework, however, topology is not directly affected by missing data \textit{per se}, but by the number of morphological characters shared among the taxa.
Therefore, the two main drivers of incorrect topologies are the overall number of morphological characters and the number of living species with no morphological data.

Our results suggest that, in order to use Total Evidence approaches, one should reduce the missing data in the morphological part of the matrix for living species and use a Maximum Likelihood framework to fix the topology prior to the overall Bayesian phylogenetic inference process.
%More results!
\end{abstract}

\noindent (Keywords: missing data, Total Evidence, Bayesian, Maximum Likelihood, topology)\\

@@

Although most species that have ever lived are now extinct % NC: This avoids giving the dodgy stat of 1%
    \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}. %or is citing of your own papers in the too "douchebagy" for the second sentence?
% NC: Yeah be careful with self citing. Fine if it's a good fit, but it's totally not here!!! A better citation would be a big biodiversity study, e.g. the bird phylogeny paper.
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}.
This has led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013}). %Not a nice sentence

Up to now, three main approaches have been used to place both living and fossil taxa % NC: BE consistent. Either living and fossil, or fossil and living. - TG: Let's make it living and fossil then.
    into phylogenies.
These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). % NC: Not sure about bit in brackets - TG: it's the idea of fossil occurence (age) only when node and fossil occurence + topology when tip.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny.
Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty (but see \citealp{spencerefficacy2013}).
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence date % NC: Need to check the exact phrase needed here. Is it range of occurrence? Occurrence dates? etc. - TG: in paleo jargon we have FAD (first apparition datum) and FOD (first occurrence datum). FAD is always hypothetical (we can't really exactly know when the taxa appeared) and FOD is always subjective (it depends when the fossil was found, maybe next year someone will find another one in older layers). So for the dating, because it's relaxed clock, people use FOD (and also probably because they don't care since it doesn't make a huge difference I guess).
    are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013}.
However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. %There might be a philosophico-semantic distinction to do here. Is TEM a method or an approach? I'll say that it is an approach because it's more like an idea (put every thing together) than a way to do (how to put everything together). It makes more sense to me compared to a real method e.g. a probabilistic method such as ML where the method states precisely how to do stuff.
% NC: Tricky. Don't think it's too important though. 
% NC: I do think Total Evidence needs capital letters though, like ML and Bayesian - TG: fixed
This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}.
Total Evidence methods have been successfully applied to empirical data \citep{pyrondivergence2011,ronquista2012,schragocombining2013}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies.
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data that can be difficult (or impossible) to collect.
%In particular it requires morphological data from both living and fossil taxa, both of which are known to be scarce. % NC: Really? I know fossil data is known to be scarce, but living data? I think people would be surprised. And they are surprised. We were surprised! Needs a citation either way.
%TG: how about this:
The morphological data for living taxa is rarely collected when molecular data is available (e.g. \citealt{O'Leary08022013} vs. \citealt{meredithimpacts2011}),% TG: i.e. in O'Leary they use meredith molecular data but only code few species from it.
 and, for fossil taxa, the scarcity of the fossil record only allow to collect the data available (for example, in vertebrates, the hardest parts of the skeleton; \citealt{sansomfossilization2013}).
Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}. % Trevor: Missing data also influences support values that trend to decrease

% NC: You need to take a bit of time to learn about singular and plural grammar in English. It's something you always do wrong. e.g. The Total Evidence method *uses* this and *its* parameters are as follows. Versus: Total Evidence methods *use* this and *their* parameters are as follows. Sive and I won't be around to correct your English forever. - TG: my French demons are back!

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013}.
Missing molecular data has been seen by some authors as an issue because it can, in some part of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing to infer topology and branch length)% NC: define phylogenetic signal here. 
    , especially when using large matrices \citep{lemmonthe2009}.
However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered genes (i.e. approximatively of the genes that are coded for most of the species; \citealt{rouresite-specific2011}) %NC: what do you mean by highly-covered?
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and
(iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}.
Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013}.
Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies \citep{sansomfossilization2013}.
However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold were it starts fucking up the phylogeny) but more the fact that Bayesian fails.

@@

% NC: Somewhere here you need to flag that you're only looking at topologies and briefly explain why.
In this study, we focused only on topology as one of the two aspects of the phylogenetic signal (topology and branch length).
Even though both aspects are equally important, branch topology is the first and most straightforward aspect reflecting phylogenetic signal (i.e. topological changes are discrete opposed to branch length changes are continuous).
Also, interestingly, the effect of Total Evidence method has not been formally assessed in previous studies using fixed topology \citep{ronquista2012,schragocombining2013}. %TG: or is that mean for them?

Here we use simulations to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
The molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}.
The effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}, therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness of the morphological part of the matrix:
\begin{enumerate}
\item the proportion of living taxa with no morphological data;
\item the proportion of missing data in the fossil taxa; and
\item the proportion of missing morphological characters for both living and fossil taxa in the matrix.
\end{enumerate}
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods.
We chose these parameters because they reflect empirical biases in data availability.
The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
Missing data in fossil taxa is very common due to preservation biases \citep{sansomfossilization2013}, and the overall number of characters depends on the effort of the people identifying them.
% NC: OK I'm not sure what the "real-life" reason for three is. - TG: fixed to "empirical" (or "practical" ?).
% TG: This paragraph is crap
% NC: I think it's fine!

We find that when using a Maximum Likelihood approach, as missing data increases, the likelihood of recovering the correct tree topology decreases.
However, even with no missing data, Total Evidence matrices dramatically reduce the performance of Bayesian methods for inferring tree topology.
We propose that this drastic difference between Bayesian and Maximum Likelihood methods is due to a flattening of the likelihood landscape caused by the unavoidable amount of missing molecular data for fossil taxa in a Total Evidence matrix.
We make suggestions for how best to deal with this issue when inferring phylogenies from Total Evidence matrices.
% NC: You don't want to tell the whole story here. Just a results summary and a quick line of interpretation

%This problem leads to an inability of Bayesian chains to pick up the sub and near optimal topologies leading to high variance in the posterior tree distribution and really low support in the consensus tree.
%These results should encourage authors in not using Bayesian inference to explore the topological landscape of their phylogeny.
%Instead, we suggest that authors should first find the Maximum Likelihood tree and use it's topology as a fix prior in their Bayesian inference.
%We also propose a new method for comparing Bayesian posterior tree distribution's topology available in a wonderful R package. %Or am I just dreaming a bit too much here? 
% NC: Dreaming a bit much I think. There's a discussion to be had here about whether it's really a *new* method and whether you really want to make an R package per se. We definitely want to share the code

@@

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------

\section{Methods} %no material needed?
% NC: Depends on the journal. See what Syst Biol asks for as section titles. TG: Syst Biol instructions :Elements vary, of course, depending on the type of paper. The first-level heading is often used for sections such as ``methods and materials,'' ``results,'' and' ``discussion;'' but other major sections are often appropriate instead.
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}). %NC: does Syst Biol use Fig. or Figure? - TG: it's "Fig." when it's a reference in brackets and "Figure" when it's a reference in the text. Table is always refered as "Table" (not "Tab").
\begin{enumerate}
\item{Generating the matrix} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the "true" tree; Table ~\ref{Tab_glossary}) and used it to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix; Table ~\ref{Tab_glossary}).
\item{Removing data} \label{step:remove_data} \\
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$) % NC: make sure to use the same descriptions here and in the introduction 
(the resulting matrices are called hereafter "missing-data" matrices; Table ~\ref{Tab_glossary}).
\item{Building phylogenies} \label{step:build_phylo} \\
We built phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the "best" tree; Table ~\ref{Tab_glossary}) and multiple trees inferred from matrices with missing morphological data (hereafter called the "missing-data" trees; Table ~\ref{Tab_glossary}). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies} \label{step:compare_topo} \\
We compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated steps 1 to 4 50 times. %that's because I had 3 cluster accounts available... If it's a too odd number, no worries, I'll just remove the last chain.
% NC: Yeah 51 is odd. Let's just use 50.
A list of all the terms used in this paper is available in Table ~\ref{Tab_glossary}.

@@

% NC: I think this section is suffering a bit from "here is everything I know about tree comparison metrics" rather than explaining what is important for your analysis. You may be able to cut some more stuff out. - TG: since I'm just using Bogdanowicz allready published code, I maybe don't need to detail the metrics at all? Just say what they measure and shovel all the detail in supp?
